---
title: "Life Expectancy from WHO Dataset"
author: "Shi Tang"
date: "10/03/2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(dplyr)
library(ggplot2)

library(data.table)
library(dtplyr)

library(sjPlot)
library(sjmisc)
library(sjlabelled)
library(ggpubr)

```

## Introduction
Provide background on your dataset and formulated question
### How does immunization coverage affect life expectancy for year 2015?
A person's life expectancy is a statistical measure of the average time is expected to live, based on several factors such as birth, age, and other demographic factors. Indeed, the United Nations estimates a global average life expectancy of 72.6 years for 2019 – which is higher than in any country back in 1950. Therefore, I am interested in exploring the impact of immunization coverage on life expectancy for the year 2015.

The data-set was from Kaggle https://www.kaggle.com/kumarajarshi/life-expectancy-who, and the original data related to life expectancy, health factors for 193 countries have been collected from WHO data repository website and its corresponding economic data was collected from United Nation website.

## Methods
include how and where the data were acquired, how you cleaned and wrangled the data, what tools you used for data exploration

### Missing values
```{r import data, echo = FALSE, include = FALSE}
data <- read_csv("data/life_expectancy_data.csv")
data <- data %>%
  rename("Life_expectancy" = "Life expectancy", "Adult_Mortality" = "Adult Mortality", "infant_deaths" = "infant deaths", "percentage_expenditure" = "percentage expenditure", "Hepatitis_B" = "Hepatitis B", "under_five_deaths" = "under-five deaths", "Total_expenditure" = "Total expenditure", "thinness_1_19_years" = "thinness  1-19 years", "thinness_5_9_years" = "thinness 5-9 years", "Income_composition_of_resources" = "Income composition of resources", "HIV_or_AID" = "HIV/AIDS")
```

```{r}
null_values <- data %>% summarize_all(funs(sum(is.na(.))))
null_values <- gather(null_values, Factors, 'missing (count)', Country:Schooling, factor_key=TRUE)
null_values['missing (%)'] <- round(null_values['missing (count)']/dim(data)[1] * 100, 3)

knitr::kable(null_values, digits = 4)
```

The data-set consists of 22 columns and 2938 rows. 

```{r checking NAs, echo = FALSE, include = FALSE}
NA_columns <- sapply(data, function(x) sum(is.na(x)))
names(NA_columns[NA_columns > 0])

for (i in names(NA_columns[NA_columns > 0])) {
  data[, i][is.na(data[, i])] <- as.numeric(lapply(data[, i], mean, na.rm = TRUE))
}
```

```{r}
data <- data %>%
  mutate(developed = ifelse(Status == 'Developed', 1, 0))

data <- data %>%
  select(-Status)
```

```{r}
normalize <- function(x){
  round((x-min(x))/(max(x)-min(x)), 4)
}

data[3:21] <- as.data.frame(lapply(data[3:21], normalize))
```

```{r}
hist(data$Life_expectancy, breaks = 30, xlab="Life expectancy", main = "Histogram of life expectancy")
```

Life Expectancy is normally distributed and is thus fit for prediction using linear regression


```{r warning=FALSE, message=FALSE, fig.height=12, fig.width=10}
p1 <- ggplot(data = data, aes(x = Adult_Mortality, y = Life_expectancy)) + 
  geom_point(color='blue') +
  geom_smooth(method = "lm", se = FALSE)

mod1 <- lm(Life_expectancy ~ Adult_Mortality, data = data)

p2 <- ggplot(data = data, aes(x = infant_deaths, y = Life_expectancy)) + 
  geom_point(color='blue') +
  geom_smooth(method = "lm", se = FALSE)

mod2 <- lm(Life_expectancy ~ infant_deaths, data = data)
  
p3 <- ggplot(data = data, aes(x = Alcohol, y = Life_expectancy)) + 
  geom_point(color='blue') +
  geom_smooth(method = "lm", se = FALSE)

mod3 <- lm(Life_expectancy ~ Alcohol, data = data)

p4 <- ggplot(data = data, aes(x = percentage_expenditure, y = Life_expectancy)) + 
  geom_point(color='blue') +
  geom_smooth(method = "lm", se = FALSE)

mod4 <- lm(Life_expectancy ~ percentage_expenditure, data = data)

p5 <- ggplot(data = data, aes(x = Hepatitis_B, y = Life_expectancy)) + 
  geom_point(color='blue') +
  geom_smooth(method = "lm", se = FALSE)

mod5 <- lm(Life_expectancy ~ Hepatitis_B, data = data)

p6 <- ggplot(data = data, aes(x = Measles, y = Life_expectancy)) + 
  geom_point(color='blue') +
  geom_smooth(method = "lm", se = FALSE)

mod6 <- lm(Life_expectancy ~ Measles, data = data)

p7 <- ggplot(data = data, aes(x = BMI, y = Life_expectancy)) + 
  geom_point(color='blue') +
  geom_smooth(method = "lm", se = FALSE)

mod7 <- lm(Life_expectancy ~ BMI, data = data)

p8 <- ggplot(data = data, aes(x = under_five_deaths, y = Life_expectancy)) + 
  geom_point(color='blue') +
  geom_smooth(method = "lm", se = FALSE)

mod8 <- lm(Life_expectancy ~ under_five_deaths, data = data)

p9 <- ggplot(data = data, aes(x = Polio, y = Life_expectancy)) + 
  geom_point(color='blue') +
  geom_smooth(method = "lm", se = FALSE)

mod9 <- lm(Life_expectancy ~ Polio, data = data)

p10 <- ggplot(data = data, aes(x = Total_expenditure, y = Life_expectancy)) + 
  geom_point(color='blue') +
  geom_smooth(method = "lm", se = FALSE)

mod10 <- lm(Life_expectancy ~ Total_expenditure, data = data)

p11 <- ggplot(data = data, aes(x = Diphtheria, y = Life_expectancy)) + 
  geom_point(color='blue') +
  geom_smooth(method = "lm", se = FALSE)

mod11 <- lm(Life_expectancy ~ Diphtheria, data = data)

p12 <- ggplot(data = data, aes(x = HIV_or_AID, y = Life_expectancy)) + 
  geom_point(color='blue') +
  geom_smooth(method = "lm", se = FALSE)

mod12 <- lm(Life_expectancy ~ HIV_or_AID, data = data)

p13 <- ggplot(data = data, aes(x = GDP, y = Life_expectancy)) + 
  geom_point(color='blue') +
  geom_smooth(method = "lm", se = FALSE)

mod13 <- lm(Life_expectancy ~ GDP, data = data)

p14 <- ggplot(data = data, aes(x = Population, y = Life_expectancy)) + 
  geom_point(color='blue') +
  geom_smooth(method = "lm", se = FALSE)

mod14 <- lm(Life_expectancy ~ Population, data = data)

p15 <- ggplot(data = data, aes(x = thinness_1_19_years, y = Life_expectancy)) + 
  geom_point(color='blue') +
  geom_smooth(method = "lm", se = FALSE)

mod15 <- lm(Life_expectancy ~ thinness_1_19_years, data = data)

p16 <- ggplot(data = data, aes(x = thinness_5_9_years, y = Life_expectancy)) + 
  geom_point(color='blue') +
  geom_smooth(method = "lm", se = FALSE)

mod16 <- lm(Life_expectancy ~ thinness_5_9_years, data = data)

p17 <- ggplot(data = data, aes(x = Income_composition_of_resources, y = Life_expectancy)) + 
  geom_point(color='blue') +
  geom_smooth(method = "lm", se = FALSE)

mod17 <- lm(Life_expectancy ~ Income_composition_of_resources, data = data)

p18 <- ggplot(data = data, aes(x = Schooling, y = Life_expectancy)) + 
  geom_point(color='blue') +
  geom_smooth(method = "lm", se = FALSE)

mod18 <- lm(Life_expectancy ~ Schooling, data = data)

p19 <- ggplot(data = data, aes(x = developed, y = Life_expectancy)) + 
  geom_point(color='blue') +
  geom_smooth(method = "lm", se = FALSE)

mod19 <- lm(Life_expectancy ~ developed, data = data)

ggarrange(p1, p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, p12, p13, p14, p15, p16, p17, p18, p19,
          labels = c("A", "B", "C", "D", "E", "F", "G", "H", "I", "J", "K", "L", "M", "N", "O", "P", "Q", "R", "S"),
          ncol = 4, nrow = 5)
```


```{r}
adjusted_R_squared <- c(0.4847, 0.03827, 0.1531, 0.1455, 0.0412, 0.0245, 0.3125, 0.04918, 0.2128, 0.04293, 0.2258, 0.3094, 0.185, 4.55e-05, 0.2227, 0.2175, 0.4793, 0.5111, 0.232)

variables <- c('Adult_Mortality', 'infant_deaths', 'Alcohol', 'percentage_expenditure', 'Hepatitis_B', 'Measles', 'BMI', 'under_five_deaths', 'Polio', 'Total_expenditure', 'Diphtheria', 'HIV_or_AID', 'GDP', 'Population', 'thinness_1_19_years', 'thinness_5_9_years', 'Income_composition_of_resources', 'Schooling', 'developed')

table <- cbind(Variable = variables, 'Adjusted R Squared' = adjusted_R_squared)
knitr::kable(table, digits = 4)

dropped <- c('Country', 'Year', 'infant_deaths', 'Alcohol', 'percentage_expenditure', 'Hepatitis_B', 'Measles', 'under_five_deaths', 'Total_expenditure', 'GDP', 'Population')

filter_data <- data %>%
  select(-one_of(dropped))

filter_data
```


```{r}
mod <- lm(Life_expectancy~., data = filter_data)
summary(mod)
```

```{r}
filter_data <- filter_data %>%
  select(-c(thinness_1_19_years, thinness_5_9_years))

mod_ <- lm(Life_expectancy~., data = filter_data)
summary(mod_)
```

```{r}
set.seed(101)
sample <- sample.int(n = nrow(filter_data), size = floor(.75*nrow(filter_data)), replace = F)
train <- filter_data[sample, ]
test  <- filter_data[-sample, ]

mod_ <- lm(Life_expectancy~., data = train)
summary(mod_)

pred_test <- predict(mod_, test)
pred_data <- data.frame(pred = pred_test, actual = test$Life_expectancy)
head(pred_data)

mean((pred_data$actual - pred_data$pred)^2)
```

## Results
provide final, publication ready tables and figures from your analysis, refer to your website if needed

## Conclusions and Summary
To improve the Life Expectancy:

Polio, Hepatitis, Diptheria vaccination coverage should be increased

Measures should be taken to ensure food security

Measures should be taken to provide education and reduce the risks of infant mortality

Resources should be utilized productively

AIDS awarness campaigns should be organized.